\section{Equivalence to First-Order Logic}

Nested conditions can express precisely the same properties of graphs as First-Order Logic (FOL). This in fact follows from previous work, but since later on in this paper we strengthen the result to the existence of functors between the respective categories, we start out by recalling the relevant definitions and giving a direct proof of the equivalence. This will also help in developing an intuition about nested conditions.

We restrict to binary predicates $\la,\lb$, taken from a set $\Lab$; furthermore, we use variables $x,y$ taken from a set $\Var$. The syntax of FOL that we use is given by the grammar
%
\[ \phi \:::=\: \True
        \:\mid\: \False
		\:\mid\: \la(x_1,x_2)
        \:\mid\: x_1\Eq x_2
		\:\mid\: \phi_1\wedge \phi_2
		\:\mid\: \phi_1\vee \phi_2
		\:\mid\: \neg\phi_1
		\:\mid\: \exists X\st \phi_1 
		\]
where $\la\in\Lab$ is an arbitrary predicate name, $x_1,x_2\in\Var$ are arbitrary variables and $X\subseteq \Var$ an arbitrary set of variables. We use the common $\exists x\st \phi$ as syntactic sugar for $\exists\setof x\st \phi$. We also use the concept of \emph{free variables} of $\phi$, denoted $\fv(\phi)$, inductively defined in the usual way.

In this section we use the following concrete category of graphs.
%
\begin{definition}
An edge-labelled multi-graph is a tuple $\tupof{N,E,s,t,\ell}$ where
\begin{itemize}
\item $N$ is a set of nodes;
\item $E$ is a set of edges;
\item $s,t\of E\to N$ are the source and target function, respectively;
\item $\ell\of E\to \Lab$ is the edge labelling function.
\end{itemize}
A multi-graph morphism $f\of A\to B$ is a tuple $f=\tupof{f\N,f\E}$ where $f\N\of N_A\to N_B$ and $f\E\of E_A\to E_B$ are functions satisfying $f\E;s_B=s_A;f\N$, $f\E;t_B=t_A;f\N$ and $f\E;\ell_B=\ell_A$.
\end{definition}
%
Although our theory mostly abstracts from node and edge identities by relating graphs through morphisms, in the context of FOL interpretation we sometimes have to rely on meaningful, fixed node identities (namely where they correspond to variables). We call $A$ and $B$ \emph{consistent} if their source, target and label functions coincide on $E_A\cap E_B$, and \emph{disjoint} if $N_A\cap N_B=E_A\cap E_B=\emptyset$ (in which case they are certainly consistent).  If $A$ and $B$ are disjoint, we use $A\cup B$ to denote their union; also, if $g\of A\to G$ and $h\of B\to G$ for disjoint $A$ and $B$, then $g\cup h\of (A\cup B)\to G$ denotes the union of $g$ and $h$. We call $A$ a subgraph of $B$, denoted $A\subseteq B$, if $B=A\cup B$; likewise, $f$ is a sub-morphism of $g$, denoted $f\subseteq g$, if $f\cup g=g$. If $A$ and $B$ are consistent, then $B\setminus A$ denotes the largest subgraph of $B$ with node set $N_B\setminus N_A$, and for $g\of A\to G$ we use $g\restr B$ and $g\setminus B$ to denote the restriction of $g$ to (the nodes and edges of) $B$ and $A\setminus B$, respectively.\todo{We use $g\setminus B$ also if $B$ is not a subgraph of $A$, e.g., in the definition of $g\sat \exists E\st \phi_1$. I think it is actually well-defined.} Finally, if $X$ is a set then $\tupof X$ will denote the discrete (i.e., edge-less) graph with node set $X$.

For the correspondence of FOL to nested conditions, we equate $\Var$ to the universe of node identities. A graph morphism $g\of A\to G$ can then be seen as an assignment of free variables (the nodes of $A$) to a given domain (the graph $G$) in which the existence of a $\la$-labelled edge between $g(x)$ and $g(y)$ means that the predicate $\la(x,y)$ holds. Formally, satisfaction is captured by a relation $\sat$ inductively defined over the structure of the formula. In fact, for reasons to become clear below, we defined $g\sat \phi$ for arbitrary $g\of A\to G$ such that $\fv(\phi)\subseteq N_A$, without requiring that $A$ is discrete.
%
\[\begin{array}{ll}
g\sat \True & \text{always} \\
g\sat \False & \text{never} \\
g\sat \la(x_1,x_2) & \iffdef \text{there is an } e\in E_G \text{ with } s(e)=g(x_1), \ell(e)=\la, t(e)=g(x_2) \\
g\sat x_1\Eq x_2 & \iffdef g(x_1)=g(x_2) \\
g\sat \phi_1\wedge \phi_2 & \iffdef g\sat \phi_1 \text{ and } g\sat \phi_2 \\
g\sat \phi_1\vee \phi_2 & \iffdef g\sat \phi_1 \text{ or } g\sat \phi_2 \\
g\sat \neg\phi_1 & \iffdef \text{not } g\sat \phi_1 \\
g\sat \exists X\st \phi_1  & \iffdef h\sat \phi_1 \text{ for some } h \text{ with } h\setminus \tupof X =g\setminus \tupof X \enspace.
\end{array}\]
%
When comparing the satisfaction relations for nested conditions and FOL formulas, a technical issue is that the sets of models for which they are defined are not the same: as seen above, $g\sat\phi$ is defined for $g\of A\to G$ whenever $\fv(\phi)\subseteq N_A$, whereas $g\sat\cT$ for a nested condition $\cT$ is only defined if $A$ is \emph{precisely} $I_\cT$. Due to this discrepancy, there cannot be any $\phi$ and $\cT$ such that $g\sat \phi$ if and only if $g\sat \cT$; instead, the best we can hope for is that this holds whenever $A=\tupof{\fv(\phi)}$. The following property implies that this is indeed all we need to be interested in, since $g$-images for $A$-elements outside $\fv(\phi)$ do not make a difference for satisfaction.

\begin{proposition}\label{prop:free vars only}
Let $\phi$ be a FOL-formula and $g\of A\to G$ such that $\fv(\phi)\subseteq N_A$; then $g\sat \phi$ if and only if $g\restr \tupof{\fv(\phi)}\sat \phi$.
\end{proposition}
%
Because of this property, when constructing a morphism $g\of A\to G$ such that $g\sat \phi$, we may assume $A$ to be any graph for which $\tupof{\fv(\phi)}\subseteq A$, without loss of generality.

In the remainder of this section, we show that
%
\begin{inumerate}
\item for every nested condition, there is an essentially equivalent FOL formula, and
\item for every FOL formula, there is an essentially equivalent nested condition.
\end{inumerate}
%
\subsection{From nested conditions to FOL}

We first define formulas for graphs, and then (inductively) for nested conditions. Let $P$ be an arbitrary graph.
%
\[ \phi_P = \textstyle
% \bigwedge_{x\in N_P}{x=x} \wedge
 \bigwedge_{e\in E_P} \ell(e)\bigl(s(e),t(e)\bigr)
\]
%
Note that $\fv(\phi_P)=N_P$. The essential property of $\phi_P$ is the following.
%
\begin{proposition}\label{prop:graph formula}
Let $P$ be an arbitrary graph.
\begin{enumerate}
\item If $g\of P\to G$ is an arbitrary graph morphism, then $g\sat \phi_P$;
\item If $g\of \tupof{N_P}\to G$ such that $g\sat \phi_P$, then $g\subseteq g'$ for some $g'\of P\to G$.
\end{enumerate}
\end{proposition}
%
In the definition below, we assume (without loss of generality, because we can always rename node identities up to isomorphism) that for all nodes $v$ of a nested condition, $U_v$ and $P_v$ are disjoint; in other words, on each successive layer of the condition, the node identities (here treated as variables) are distinct. 

Let $\cT$ be an arbitrary nested condition. We define $\psi_v,\phi_v$ for all nodes $v\in V_\cT$, inductively on the depth of the subtree $\hat v$:
%
\begin{align}
\psi_v = {} & \textstyle \phi_{P_v}\wedge \bigwedge_{x\in N_{I_v}} d_v(x)\Eq u_v(x) \wedge \bigwedge_{w\childof v} \neg\phi_w \label{eq:psi} \\
\phi_v = {} & \exists N_{P_v}\st \psi_v
\end{align}
%
Finally, we define $\phi_\cT=\phi_{\rtof\cT}$.
%
\begin{theorem}
For an arbitrary nested tree $\cT$ and an arbitrary open graph $g\of I_\cT\to G$, $g\sat\phi_\cT$ if and only if $g\sat\cT$.
\end{theorem}
%
\begin{proof}
For our proof, we need an induction hypothesis for all nodes $v$ of a nested condition, in which the $u_v$-morphism is also accounted for. The property we actually prove is therefore:
\begin{quote}
For all nodes $v$ of $\cT$ and all $g\of U_v\to G$, $g\sat \phi_v$ if and only if $u_v;g\sat \hat v$.
\end{quote}
This is proved inductively: we assume that the property holds for all $w\childof v$, and show that it then holds for $v$. Since $u_{\rtof\cT}=\id_{I_\cT}$ and hence $U_{\rtof\cT}=I_\cT$, this implies the proof obligation.
%
\begin{description}
\item[If.] Assume $u_v;g\sat \hat v$, meaning that there is some $h\of P_v\to G$ such that $d_v;h=u_v;g$ and $u_w;h\nsat \hat w$ for all $w\childof v$. Since (by assumption) $P_v$ and $U_v$ are disjoint, we can construct the graph $A=P_v\cup U_v$ and the graph morphism $k=g\cup h\of A\to G$. Note that $P_v=A\setminus N_{P_v}$ and hence $g=k\setminus N_{P_v}$.

\smallskip
By \Cref{prop:graph formula}, it follows that $h\sat \phi_{P_v}$. Moreover, by the induction hypothesis, $h\nsat \phi_w$, implying $h\sat\neg \phi_w$. \Cref{prop:free vars only} then implies $k\sat \phi_{P_v}$ and $k\sat \neg\phi_w$. Also, for any $x\in N_{I_v}$,
\[ k(d_v(x))=g(d_v(x))=(d_v;g)(x) = (u_v;h)(x) = h(u_v(x))=k(u_v(x)) \enspace, \]
hence $k\sat d_v(x)\Eq u_v(x)$. All in all, we have $k\sat \psi_v$. Since $g\setminus \tupof{N_{P_v}}= g= \tupof{k\setminus N_{P_v}}$, it follows that $g\sat \phi_v$.

\smallskip
\item[Only if.] Assume $g\sat \phi_v$. This means that there is a $k\of A\to G$ such that $g\setminus \tupof{N_{P_v}}=k\setminus \tupof{N_{P_v}}$ and $k\sat \psi_v$ (see \Cref{eq:psi}). Because $P_v$ and $U_v$ are disjoint, due to \Cref{prop:free vars only} we may assume without loss of generality that $A= U_v\cup\tupof{N_{P_v}}$, and hence $k=g\cup h$ for some $h\of \tupof{N_{P_v}}\to G$. Based on this, we can show the existence of a witness for $g\sat \hat v$, as follows.
\begin{itemize}
\item $k\sat \phi_{P_v}$ implies (by \Cref{prop:free vars only}) $h\sat \phi_{P_v}$, and hence (by \Cref{prop:graph formula}) $h\subseteq h'$ for some $h'\of P_v\to G$. \Cref{prop:free vars only} then implies $h'\sat \phi_{P_v}$.

\item Let $k'=g\cup h'$; then $k'\supseteq k$ and hence (again by \Cref{prop:free vars only}) $k'\sat \psi_v$.

\item $k'\sat \bigwedge_{x\in N_{I_v}} d_v(x)\Eq u_v(x)$ means that $h'(d_v(x))=k'(d_v(x))=k'(u_v(x))=g(u_v(x))$ for all $x\in N_{I_v}$, and since $I_v=\tupof{N_{I_v}}$ this implies $d_v;h'=u_v;g$.

\item $k'\sat \neg\phi_w$ (for all $w\childof v$) implies (by \Cref{prop:free vars only}) $h'\sat\neg\phi_w$, hence $h'\nsat\phi_w$, hence (by the induction hypothesis) $u_w;h'\nsat \hat w$.
\end{itemize}

Taken together, this establishes that $h'$ is a witness for  $g\sat \phi_v$.
\end{description}
\end{proof}
%
The formula $\phi_\cT$ can in practice be simplified by alpha-converting the variable names such that they are shared as much as possible, rather than being distinct on each layer as required for the definition; depending on the condition tree, this may allow many or even all of the equations $d_v(x)\Eq u_v(x)$ in $\phi_v$ to be omitted. This is in particular the case if both $d_v$ and $u_v$ are injective.

\subsection{From FOL to nested conditions}

This is shown inductively over the structure of FOL formulas. Indeed, we show that
\begin{itemize}
\item for every basic predicate in FOL, there exists a corresponding nested condition;
\item for every operator in FOL there exists a construction over nested conditions that mimics it.
\end{itemize}
%
Because of the De Morgan duality $\phi\vee \phi\equiv \neg(\neg\phi\wedge \neg\psi)$, in fact we only have to give constructions (of the second kind) for $\wedge$, $\neg$ and $\exists$.

\begin{itemize}
\item $\cT_\True$ is given by $V=\setof v$ with $I_v=P_v=\tupof\emptyset$ (the empty graph), where $u_v=d_v=\id_{I_v}$.
\item $\cT_\False$ is given by $V=\setof{v,w}$ with $v\parentof w$, $I_v=P_v=I_w=P_w=\tupof\emptyset$, where $u_v=d_v=u_w=d_w=\id_{I_v}$.
\item $\cT_{\la(x_1,x_2)}$ is given by $V=\setof{v}$ with $I_v=\tupof{\setof{x_1,x_2}}$ and $P_v\supset I_v$ having a single $\la$-labelled edge from $x_1$ to $x_2$, where $u_v=\id_{I_v}$ and $d_v\of I_v\to P_v$ is the morphism that acts as the identity on the $\setof{x_1,x_2}$.
\item $\cT_{x_1\Eq x_2}$ is given by $V=\setof v$ with $I_v=\tupof{\setof{x_1,x_2}}$ and $P_v=\tupof{\setof{y}}$, where $u_v=\id_{I_v}$ and $d_v\of I_v\to P_v$ is the only available morphism.
\end{itemize}

\begin{proposition}
Let $\cT$ be a nested condition and $g\of I_\cT\to G$ an open graph.
\begin{enumerate}
\item If $\cT=\cT_\True$, then $g\sat \cT$ if and only if $g\sat\True$.
\item If $\cT=\cT_\False$, then $g\sat \cT$ if and only if $g\sat\True$.
\item If $\cT=\cT_{\la(x_1,x_2)}$, then $g\sat \cT$ if and only if $g\sat\la(x_1,x_2)$.
\item If $\cT=\cT_{x_1\Eq x_2}$, then $g\sat \cT$ if and only if $g\sat x_1\Eq x_2$.
\end{enumerate}
\end{proposition}

To encode negation of nested conditions, we introduce a construction $\neg\cT$ that essentially ``pushes'' $\cT$ one level down while introducing a new root, which itself imposes no structural constraints. $\cU=\neg\cT$ is given by $V_\cU=\tupof{V_\cT\cup\setof w}$ with ${\parentof_\cU}= {\parentof_\cT}\cup \setof{(w,\rtof\cT)}$, $I_w=P_w=I_\cT$ and $d_v=u_v=\id_{I_\cT}$, whereas for all $v\in V_\cT$, $d_v$ and $u_v$ remain as they were in $\cT$.

\begin{proposition}
Let $\cT$ be a nested condition and $g\of I_\cT\to G$ an open graph; then $g\sat \neg\cT$ if and only if $g\nsat\cT$.
\end{proposition}

To define the conjunction of nested conditions, we first have to make the operands consistent by ensuring that they have the same interface. For this purpose, we define an auxiliary construction $\cT\otimes J$ for any discrete graph $J$ disjoint from both $I_{\rt}$ and $P_{\rt}$. $\cU=\cT\otimes J$ is equal to $\cT$ except that $J$ is ``added'' to the root; hence
\begin{itemize}
\item $I^\cU_{\rt}=I^\cT_{\rt}\cup J$ and $P^\cU_{\rt}=P^\cT_{\rt}\cup J$, with embedding $\epsilon\of P^\cT_{\rt}\to P^\cU_{\rt}$;
\item $d^\cU_{\rt}=d^\cT_{\rt}\cup \id_J$;
\item $u^\cU_w=u^\cT_w;\epsilon$ for all $w\childof \rt$. 
\end{itemize}


\begin{proposition}
Let $\cT$ be a nested condition and $J$ a discrete graph disjoint from both $I_{\rt}$ and $P_{\rt}$. For all $g\of I_{\rt}\cup J\to G$, $g\sat \cT\otimes J$ if and only if $g\setminus J\sat\cT$.
\end{proposition}

The last step is to define the \emph{product} of two nested conditions with the same interface. This is based on the pushout of the root morphisms. Let $\cT_i$ for $i=1,2$ be nested conditions with $\rt=\rt_1=\rt_2$, $V_1\cap V_2=\setof{\rt}$ and $I_{\cT_1}=I_{\cT_2}=I$, and denote $P_i=P_{i,\rt}$ and $d_i=d_{i,\rt}$ for $i=1,2$. The pushout of $d_1$ and $d_2$ consists of a pushout object $P=P_1\times_I P_2$ and morphisms $d'_i\of P_i\to P$ for $i=1,2$, and for every morphism $f\of A\to P_i$ there is an extended morphism $f;d'_i\of A\to P$; in particular, we denote $d=d_1;d'_1=d_2;d'_2\of I\to P$. Now we define $\cU=\cT_1\times \cT_2$ with $V=V_1\cup V_2$, ${\parentof}= {\parentof}_1\cup {\parentof}_2$ and
%
\begin{itemize}
\item $I^\cU_{\rt}=I$, $P^\cU_{\rt}=P$ and $d^\cU_{\rt}=d$;
\item $u^\cU_w=u^\cP_w;d'_i$ for all $w\childof_i \rt$.
\end{itemize}

\begin{proposition}
Let $\cT_1,\cT_2$ be nested conditions with $\rt=\rt_1=\rt_2$, $V_1\cap V_2=\setof{\rt}$ and $I_{\cT_1}=I_{\cT_2}=I$. For all $g\of I\to G$, $g\sat \cT_1\times \cT_2$ if and only if $g\sat\cT_1$ and $g\sat \cT_2$.
\end{proposition}
%
The following proposition shows that we can now encode FOL conjunction.

\begin{proposition}
Let $\cT_1,\cT_2$ be nested conditions with $\rt=\rt_1=\rt_2$ and $V_1\cap V_2=\setof{\rt}$, and let $J=I_{\cT_1}\setminus I_{\cT_2}$ and $J_2=I_{\cT_2}\setminus I_{\cT_1}$. Assume that $J_1$ and $P_{\cT_2,\rt}$ as well as $J_2$ and $P_{\cT_1,\rt}$ are disjoint. For all $g\of (I_{\cT_1}\cup I_{\cT_2})\to G$, $g\sat (\cT_1\otimes J_2)\times (\cT_2\otimes J_1)$ if and only if $g\setminus J_2\sat\cT_1$ and $g\setminus J_1\sat \cT_2$.
\end{proposition}